---
title: "Differential Expression Analysis Pipeline"
author: "Chu Lab"
date: "`r format(Sys.Date())`"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
editor_options:
  chunk_output_type: console
---

```{r global-knitr-opts, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = TRUE,
  fig.width = 8, fig.height = 6
)
```

# Overview

End-to-end DESeq2 pipeline for a single pairwise contrast. Each chunk maps to a
discrete analytical step so you can step through the analysis interactively
(`Ctrl/Cmd-Shift-Enter` to run a chunk; `Ctrl/Cmd-Alt-N` for next chunk) and
inspect the intermediate objects/plots in RStudio as you go.

**Workflow:**

1. Set parameters
2. Load utility functions and libraries
3. Load + merge count data
4. Build the `DESeqDataSet`
5. Run DESeq2 + variance stabilization
6. PCA + sample correlation QC
7. Extract + annotate results
8. MA plots (overall + pathway-specific)
9. (Optional) GO/KEGG enrichment
10. (Optional) WNT pathway gene export

> **Working directory:** RStudio chunks run with the working directory set to this
> file's location. If you opened the RStudio Project at the repo root that is
> already correct. Confirm with `getwd()` if relative paths fail.

# 1. Parameters

Edit values here, then run the chunk. Everything downstream reads these.

```{r parameters}
# Analysis name (used for output directory under figures/)
ANALYSIS_NAME       <- "GroupA_vs_GroupB"

# Inputs
DATA_CONFIG         <- "config/data_sources_TEMPLATE.csv"
METADATA_FILE       <- "config/sample_metadata_TEMPLATE.txt"

# Experimental design
FACTOR_NAME         <- "CellType"    # column in metadata
BASELINE_LEVEL      <- "GroupB"      # reference
COMPARISON_LEVEL    <- "GroupA"      # test group
DESIGN_FORMULA      <- "~CellType"   # DESeq2 design

# Organism — used for Entrez mapping + enrichment
#   "human" / "Hs" / "hsa"   OR   "pig" / "Ss" / "ssc"
# Human DBs are far better annotated; pick "human" unless your gene symbols are
# strictly pig-specific. See docs/ORGANISM_SELECTION.md for details.
ENRICHMENT_ORGANISM <- "human"

# Optional steps
PERFORM_ENRICHMENT   <- TRUE   # run GO + KEGG
RUN_PATHWAY_ANALYSIS <- TRUE   # export WNT pathway subset

# Output
OUTPUT_DIR <- "figures"
```

# 2. Setup — libraries, utilities, output directory

The chunk below loads packages and sources every helper file under
`scripts/utils/`. If a package is missing, install it per the README's
"Installation" section before continuing.

```{r setup-environment}
suppressPackageStartupMessages({
  library("tidyverse")
  library("DESeq2")
  library("writexl")
  library("ggplot2")
  library("ggrepel")
  library("ggpubr")
  library("org.Hs.eg.db")
  library("org.Ss.eg.db")
  library("pheatmap")
  library("RColorBrewer")
  if (PERFORM_ENRICHMENT) {
    library("clusterProfiler")
    library("enrichplot")
  }
})

source("scripts/utils/data_processing.R")
source("scripts/utils/deseq_helpers.R")
source("scripts/utils/plotting_themes.R")
if (PERFORM_ENRICHMENT) {
  source("scripts/utils/enrichment_analysis.R")
}

analysis_dir <- file.path(OUTPUT_DIR, ANALYSIS_NAME)
dir.create(analysis_dir, showWarnings = FALSE, recursive = TRUE)

cat("Analysis:", ANALYSIS_NAME, "\n",
    "Comparison:", COMPARISON_LEVEL, "vs", BASELINE_LEVEL, "\n",
    "Output dir:", analysis_dir, "\n", sep = " ")
```

# 3. Load and merge datasets

Each row of `DATA_CONFIG` is one count file plus (optionally) a regex of
columns to drop (e.g. to strip a species suffix). After per-dataset
processing, datasets are merged on shared gene symbols.

```{r load-data}
data_config <- read.csv(DATA_CONFIG, stringsAsFactors = FALSE)
cat("Loading", nrow(data_config), "datasets\n")

dataset_list <- list()
for (i in seq_len(nrow(data_config))) {
  dataset_name <- data_config$dataset_name[i]
  filepath     <- data_config$filepath[i]
  remove_pat   <- data_config$remove_pattern[i]
  if (is.na(remove_pat) || remove_pat == "") remove_pat <- NULL

  cat("  -", dataset_name, "\n")
  dataset_list[[dataset_name]] <- process_dataset(filepath, remove_pattern = remove_pat)
}

merged_data  <- merge_datasets(dataset_list)
count_matrix <- prepare_count_matrix(merged_data)
cat("Final matrix:", nrow(count_matrix), "genes x", ncol(count_matrix), "samples\n")
```

# 4. Build the DESeqDataSet

The metadata file's rows must correspond to `colnames(count_matrix)` **in the same
order** — DESeq2 does not reorder them for you and will error if they don't line up.
`set_factor_levels` forces the reference level so the contrast direction is what you
expect.

> **Keep each comparison scoped to its two groups.** `set_factor_levels` sets the
> `FACTOR_NAME` column to exactly `c(BASELINE_LEVEL, COMPARISON_LEVEL)`; any sample
> whose value is not one of those two becomes `NA` and is silently dropped from the
> model. Configure `DATA_CONFIG` / `METADATA_FILE` so only the two groups being
> contrasted are loaded.

```{r build-dds}
dds <- create_deseq_dataset(count_matrix, METADATA_FILE, design_formula = DESIGN_FORMULA)
dds <- set_factor_levels(dds, FACTOR_NAME, c(BASELINE_LEVEL, COMPARISON_LEVEL))
dds
```

# 5. Run DESeq2 and variance stabilization

`run_deseq_analysis()` estimates size factors, dispersion, and fits the GLM.
`variance_stabilize()` returns a `vst()`-transformed object used for PCA + QC.

```{r run-deseq}
dds <- run_deseq_analysis(dds)
rld <- variance_stabilize(dds)
```

# 6. PCA and sample-correlation QC

A PCA is generated for every metadata column (so you can spot clustering by
batch, dataset, sex, etc.) and a sample correlation heatmap is exported with
its matrix. Figures display in the Plots pane and are saved as PDFs.

```{r qc-plots}
metadata_cols <- colnames(SummarizedExperiment::colData(rld))
metadata_cols <- setdiff(metadata_cols, c("sizeFactor", "replaceable"))

for (col in metadata_cols) {
  generate_pca_plot(rld,
                    file.path(analysis_dir, paste0("PCA_by_", col, ".pdf")),
                    intgroup = col)
}

generate_correlation_heatmap(rld,
                             file.path(analysis_dir, "sample_correlation_heatmap.pdf"))
```

# 7. Extract and annotate results

The Wald-test result is passed through `apeglm` LFC shrinkage, then annotated
with gene symbols + Entrez IDs for downstream enrichment. Significant hits
are written to a separate Excel file for convenience.

```{r results}
contrast_name <- paste0(FACTOR_NAME, "_",
                        make.names(COMPARISON_LEVEL), "_vs_",
                        make.names(BASELINE_LEVEL))
res <- results(dds,
               contrast = c(FACTOR_NAME, COMPARISON_LEVEL, BASELINE_LEVEL),
               alpha = 0.05)
res <- DESeq2::lfcShrink(dds, coef = contrast_name, type = "apeglm")

gene_symbols <- rownames(count_matrix)
res    <- annotate_results(res, gene_symbols, organism = ENRICHMENT_ORGANISM)
res    <- merge_counts_with_results(res, dds)
res_df <- as.data.frame(res)

cat("Significant genes (padj < 0.05):",
    sum(res_df$padj < 0.05, na.rm = TRUE), "\n")

write_xlsx(res_df,                         file.path(analysis_dir, "allGenes.xlsx"))
write_xlsx(filter_significant_genes(res_df), file.path(analysis_dir, "sigGenes.xlsx"))

nm_counts_df <- as.data.frame(counts(dds, normalized = TRUE))
nm_counts_df <- cbind("GeneID" = rownames(nm_counts_df), nm_counts_df)
rownames(nm_counts_df) <- NULL
write_xlsx(nm_counts_df, file.path(analysis_dir, "normalized_counts.xlsx"))
```

# 8. MA plots

One generic MA plot plus four pathway-highlighted variants (WNT ligands, WNT
receptors, pluripotency markers, broader WNT-linked set). The labeled plot uses
`ggrepel` for non-overlapping labels and is best for figures.

```{r ma-plots}
gene_sets <- get_gene_sets()

generate_ma_plot(res_df,
                 file.path(analysis_dir, "MAplot.pdf"),
                 title = paste(COMPARISON_LEVEL, "vs", BASELINE_LEVEL),
                 top = 20)

generate_ma_plot(res_df,
                 file.path(analysis_dir, "MAplot_WNTligands.pdf"),
                 title = "WNT Ligands",
                 highlight_genes = gene_sets$wnt_ligands)

generate_ma_plot(res_df,
                 file.path(analysis_dir, "MAplot_WNTreceptors.pdf"),
                 title = "WNT Receptors",
                 highlight_genes = gene_sets$wnt_receptors)

generate_ma_plot(res_df,
                 file.path(analysis_dir, "MAplot_Pluripotency.pdf"),
                 title = "Pluripotency Genes",
                 highlight_genes = gene_sets$pluripotency)

generate_ma_plot_labeled(res_df,
                         file.path(analysis_dir, "MAplot_WNTlinked.pdf"),
                         title = "WNT Pathway Genes",
                         highlight_genes = gene_sets$wnt_linked)
```

# 9. Enrichment analysis (optional)

`compareCluster` is run separately on Up- and Down-regulated genes (LFC > 1,
padj < 0.05) for GO/BP and KEGG. If a category has no enriched terms the
pipeline logs a message and skips the corresponding output instead of crashing.

```{r enrichment, eval=PERFORM_ENRICHMENT}
gene_lists <- get_regulated_genes(res_df, lfc_cutoff = 1, padj_cutoff = 0.05)
cat("Upregulated:",   length(gene_lists$Upregulated),   "\n")
cat("Downregulated:", length(gene_lists$Downregulated), "\n")

enrichment_results <- run_enrichment_analysis(
  gene_lists = gene_lists,
  universe   = res_df$entrezid,
  organism   = ENRICHMENT_ORGANISM,
  ontology   = "BP"
)

if (!is.null(enrichment_results$GO)) {
  write_xlsx(enrichment_results$GO@compareClusterResult,
             file.path(analysis_dir, "GO_combined.xlsx"))
}
if (!is.null(enrichment_results$KEGG)) {
  write_xlsx(enrichment_results$KEGG@compareClusterResult,
             file.path(analysis_dir, "KEGG_combined.xlsx"))
}

generate_enrichment_dotplot(enrichment_results$GO,
                            file.path(analysis_dir, "GO_dotplot.pdf"),
                            title = "GO Enrichment")
generate_enrichment_dotplot(enrichment_results$KEGG,
                            file.path(analysis_dir, "KEGG_dotplot.pdf"),
                            title = "KEGG Enrichment")
```

# 10. WNT pathway gene export (optional)

Filters `res_df` to the canonical WNT signaling gene list and writes it out
for downstream pathway-focused figures (e.g. the heatmap script).

```{r pathway-export, eval=RUN_PATHWAY_ANALYSIS}
wnt_file <- "config/WntSignalingPathway.txt"
if (file.exists(wnt_file)) {
  wnt_results <- filter_pathway_genes(wnt_file, res_df, gene_col = "GeneID")
  write_xlsx(wnt_results, file.path(analysis_dir, "WntGenes.xlsx"))
  cat("WNT pathway genes exported\n")
}
```

# Outputs

```{r summary, echo=FALSE, results='asis'}
files <- list.files(analysis_dir, full.names = FALSE)
cat("**Files in `", analysis_dir, "`:**\n\n", sep = "")
cat(paste0("- `", files, "`"), sep = "\n")
```
