---
title: "Heatmap Generator"
author: "Chu Lab"
date: "`r format(Sys.Date())`"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
editor_options:
  chunk_output_type: console
---

```{r global-knitr-opts, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = TRUE,
  fig.width = 10, fig.height = 8
)
```

# Overview

Build a single heatmap with flexible gene and sample selection from a
DESeq2-normalized count matrix (i.e. the `normalized_counts.xlsx` that
`run_differential_expression.Rmd` writes out).

You pick exactly **one** gene-selection method:

1. **All genes** (`USE_ALL_GENES`) — only useful for very small matrices.
2. **Pathway file** (`USE_PATHWAY_GENES`) — e.g. `config/WntSignalingPathway.txt`.
3. **External gene list file** (`USE_GENE_LIST_FILE`) — one gene per line.
4. **Inline custom vector** (`USE_CUSTOM_GENES`) — quick ad-hoc plots.

Samples can optionally be subset by metadata column (`SAMPLE_CRITERIA`).

# 1. Parameters

```{r parameters}
ANALYSIS_NAME <- "my_heatmap"
OUTPUT_DIR    <- "figures"

# ---- Inputs --------------------------------------------------------
COUNTS_FILE   <- "figures/my_analysis/normalized_counts.xlsx"
METADATA_FILE <- "config/sample_metadata.txt"

# ---- Gene selection (set exactly one to TRUE) ----------------------
USE_ALL_GENES       <- FALSE
USE_PATHWAY_GENES   <- TRUE
USE_GENE_LIST_FILE  <- FALSE
USE_CUSTOM_GENES    <- FALSE

PATHWAY_FILE   <- "config/WntSignalingPathway.txt"
GENE_LIST_FILE <- "path/to/gene_list.txt"
CUSTOM_GENES   <- c("WNT1", "WNT3A", "CTNNB1", "AXIN2")

# ---- Sample selection ---------------------------------------------
# NULL = use all samples. Otherwise a named list of column -> allowed values:
#   list(CellType = "GroupA")
#   list(Dataset  = c("Dataset1", "Dataset2"))
SAMPLE_CRITERIA <- NULL

# ---- Heatmap parameters -------------------------------------------
CLUSTER_ROWS    <- TRUE
CLUSTER_COLS    <- FALSE
SORT_ROWS_ALPHA <- FALSE               # TRUE orders gene rows A->Z (overrides row clustering)
COLOR_SCHEME    <- "viridis"           # "viridis" / "blue_red" / "blue_yellow"
SCALE_METHOD    <- "row"               # "0to1" / "row" (z-score) / "column" / "none"
ANNOTATION_COLS <- c("CellType", "Dataset")

PLOT_WIDTH  <- 12
PLOT_HEIGHT <- 10
```

# 2. Setup

```{r setup-environment}
suppressPackageStartupMessages({
  library("tidyverse")
  library("readxl")
  library("pheatmap")
  library("viridis")
  library("RColorBrewer")
})

source("scripts/utils/heatmap_helpers.R")
source("scripts/utils/gene_set_operations.R")

dir.create(OUTPUT_DIR, showWarnings = FALSE, recursive = TRUE)
```

# 3. Load counts and metadata

```{r load-data}
counts   <- load_normalized_counts(COUNTS_FILE)
metadata <- read.table(METADATA_FILE,
                       header = TRUE, sep = "\t",
                       row.names = 1, stringsAsFactors = FALSE)
cat("Loaded metadata for", nrow(metadata), "samples\n")
```

# 4. Select genes

The four flags are checked in order; the first `TRUE` wins.

```{r select-genes}
if (USE_ALL_GENES) {
  gene_list <- NULL
  cat("Using all genes\n")

} else if (USE_PATHWAY_GENES && file.exists(PATHWAY_FILE)) {
  gene_list <- load_pathway_genes(PATHWAY_FILE)
  cat("Using pathway genes:", length(gene_list), "\n")

} else if (USE_GENE_LIST_FILE && file.exists(GENE_LIST_FILE)) {
  gene_list <- read.table(GENE_LIST_FILE, stringsAsFactors = FALSE)[[1]]
  cat("Loaded gene list from file:", length(gene_list), "genes\n")

} else if (USE_CUSTOM_GENES) {
  gene_list <- CUSTOM_GENES
  cat("Using custom gene list:", length(gene_list), "genes\n")

} else {
  stop("No valid gene-selection method enabled. Set exactly one USE_* flag to TRUE.")
}
```

# 5. Select samples

```{r select-samples}
if (!is.null(SAMPLE_CRITERIA)) {
  sample_list <- get_samples_by_criteria(metadata, SAMPLE_CRITERIA)
} else {
  sample_list <- NULL
  cat("Using all samples\n")
}
```

# 6. Generate the heatmap

Renders to the Plots pane and writes a PDF to `OUTPUT_DIR/ANALYSIS_NAME.pdf`.

- `SCALE_METHOD = "row"` (z-score per gene) is the typical choice for comparing
  expression patterns across samples; `"0to1"` is better when you want absolute
  values bounded.
- Turn `CLUSTER_COLS` to `TRUE` to let samples self-cluster — useful for QC,
  less so when you want to enforce a specific ordering.
- `SORT_ROWS_ALPHA = TRUE` orders gene rows alphabetically (A->Z). It overrides
  row clustering, so `CLUSTER_ROWS` is ignored while it is on.

```{r heatmap}
output_file <- file.path(OUTPUT_DIR, paste0(ANALYSIS_NAME, ".pdf"))

generate_flexible_heatmap(
  count_data     = counts,
  gene_list      = gene_list,
  sample_list    = sample_list,
  metadata       = metadata,
  output_file    = output_file,
  title          = ANALYSIS_NAME,
  cluster_rows    = CLUSTER_ROWS,
  cluster_cols    = CLUSTER_COLS,
  sort_rows_alpha = SORT_ROWS_ALPHA,
  color_scheme   = COLOR_SCHEME,
  scale_method   = SCALE_METHOD,
  annotation_col = ANNOTATION_COLS,
  width          = PLOT_WIDTH,
  height         = PLOT_HEIGHT
)

cat("Heatmap PDF saved to:", output_file, "\n")
```
